Adaptive Safe Strong Screening for Efficient Lasso-Type Problems Optimization over the Solution Path

### 2021-09-08 appendix.tex
Edits to appendix

\appendix
\appendixpage

\section{Supplementary Real Data Analysis Results}
\label{sec:ap}

At the first replication on each of the 4 data sets, we first record the number of features discarded by each screening method at each $\lambda$ value along the path of $L=100$ values. Features discarded by SEDPP are safely discarded and features discarded by SSR are not safely discarded and require additional checks. For the adaptive hybrid rule, we use AHR\_Safe to denote the features that are safely discarded first by adaptive EDPP and AHR\_Strong to denote features that discarded afterward by adaptive SSR and require additional checks. The results are summarized in Figure \ref{fig:ap1}.

@@

Note each jump of the number of discards for AHR method indicates an update of reference because a newer reference leads to higher screening power. Our adaptive updating algorithm determines the timing of update dynamically depending on the screening performance. For all four data sets, with about only 10 updates, the adaptive hybrid screening can safely discard a great number of features at most $\lambda$ values.

Next we compare the number of rounds of KKT conditions checks and the number of KKT conditions checks on individual features, needed for AHR and SSR respectively, as shown in Table \ref{Tab:kkt1} and Table \ref{Tab:kkt2}. In a round of KKT conditions check, all features that are not discarded by the safe rule are checked, and a round a KKT conditions check potentially include a large number of KKT conditions checks on individual features.

@@

Although a path include $L=100$ $\lambda$ values, the solution at the first value $\lambda_{0}$ is trivially $\hat{\beta}(\lambda_0)=0$ so we are essentially solving along $99$ $\lam$ values. If the strong rule does not make a violation, the minimal number of rounds of KKT conditions checks required will be $99$. In all four data sets, neither SSR nor the adaptive SSR in AHR ever makes a violation. However two methods are very different in terms of the number of KKT conditions checks on individual features, which proportionally increases the computational cost. AHR requires much less checks on individual features because the adaptive EDPP narrows down the set of features that need to be checked by a magnitude.

>>>>

%\appendix
%\appendixpage

\section*{Appendix}
%\label{sec:ap}

To illustrate the inner workings of the proposed algorithm on a variety of problems, for a single replication of each of the four data sets in Section~\ref{sec:real-data} we recorded the number of features discarded by each screening method at each $\lambda$ value along the path of $L=100$ values. Features discarded by SEDPP are safely discarded, while features discarded by SSR are not necessarily safe and require additional checks. For the adaptive hybrid rule, we use AHR\_Safe to denote the features that are safely discarded first by adaptive EDPP and AHR\_Strong to denote features that discarded afterward by adaptive SSR and require additional checks. The results are summarized in Figure \ref{fig:ap1}.

@@

Note that the AHR method is rather ``jumpy''; this is by design, and these spikes represent moments along the path that triggered an update of the reference solution to gain more screening power. Our adaptive updating algorithm determines the timing of update dynamically depending on the screening performance. For all four data sets, with about only 10 updates, the adaptive hybrid screening can safely discard a great number of features at most $\lambda$ values.

Next we compare the number of rounds of KKT condition checks and the number of KKT condition checks on individual features needed for AHR and SSR respectively, as shown in Table \ref{Tab:kkt1} and Table \ref{Tab:kkt2}. In a round of KKT condition checking, all features that are not discarded by the safe rule are checked, which potentially includes a large number of individual feature checks.

@@

Although all paths includes $L=100$ $\lambda$ values, the solution at the first value is exactly $\hat{\beta}(\lambda_0)=0$ and does not require any checking. Thus, if no violations occur in the application of the strong rule, the number of rounds of KKT condition checking required will attain its minimum value of $99$. In all four data sets, neither SSR nor the adaptive SSR in AHR is ever violated (as other authors have discussed, violations are rare but possible). However, the two methods are very different in terms of the number of individual features involved in these condition checks. AHR requires far fewer checks on individual features because the adaptive EDPP narrows down the set of features that need to be checked by an order of magnitude, which is how it achieves greater computational efficiency.

### 2021-09-08 main.tex
Edits to appendix

This manuscript provides a brief review of existing feature screening rules in Section\ref{sec:existing} as well as the following novel contributions:

\begin{itemize}
    \item A new feature screening framework (Section 3.1,3.2), which updates the reference for screening along the grid of $\lambda$ values on any combination of strong screening rules and safe screening rules, which can generate a family of more efficient and scalable screening rules, compared to previous screening rules that are either sequential or one-step.
    \item Three specific applications of this framework, for fitting lasso-penalized linear regression, lasso-penalized logistic regression and group lasso models, respectively (Section~\ref{sec:method},~\ref{sec:logis},~\ref{sec:group}).
    \item An algorithm to determine when to update to a new reference $\lambda$ value for best efficiency (Section \ref{sec:adaptive}).
    \item Simulation studies (Section~\ref{sec:sim}) and real data analysis (Section~\ref{sec:real-data}) under a variety of scenarios, including testing data sets larger than memory, which showed that the proposed algorithms outperform other approaches in all cases, with substantial speed up.
    \item The new feature screening methods for lasso-penalized linear regression and logistic regression were implemented and added as new options to the publicly accessible \textbf{R} package \textbf{biglasso} \citep{zeng2017biglasso}, which was used to carry out the analyses in this manuscript.
\end{itemize}

@@

\subsection{Adaptive Safe Rules}

@@

\subsection{Adaptive Hybrid Rules}

@@

\begin{algorithm}[H]
  \label{alg:path}
    \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Pre}{Pre-compute}\SetKwInOut{Initialize}{Initialize}
    \SetAlgoLined
    \Input{$\{x_j\}_{j=1}^p, \{x_j^Ty\}_{j=1}^p,  \lambda_{0},...,\lambda_{L}$}
    \BlankLine
    \Initialize{reference = $\lambda_0$}

@@

This complexity analysis will only consider calculations with $O(np)$ complexity. The remaining calculations, those with $O(p)$ complexity for example, are not only of lower order, but also consume less memory and thus usually do no require out-of-core computing. Calculation of $\{x_j^Ty\}_{j=1}^p$ is needed only once for the whole solution path; we ignore this as well as its computational cost is the same across all approaches. We assume that strong rule screening is always powerful in the sense that its set of non-discarded features, $\mathcal{R}_{l+b}$, satisfies $|\mathcal{R}_{l+b}|=o(p) \approx 0$, but that safe rule screening, with corresponding set $\mathcal{S}_{l+b}$, may not be. We also assume that post-convergence check will usually pass in the first iteration and thus is only needed approximately once per $\lambda$. These two assumptions are supported by experiments from both the origin paper of SSR \citep{Tibshirani2012} and many other papers studying feature screening.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{511.eps}    \caption{Comparing speed of screening methods for lasso model under different settings. Top row: computation time in second. Bottom row: relative computation time compared to AC. First column: varying number of features $p$. Second column: varying sample size $n$. Third column: varying signal strength $\beta_{\max}$. Fourth column: varying sparsity $s$.}
    \label{fig:5.1.1a}
\end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[width=0.82\textwidth]{511b.eps}    \caption{Comparing speed of screening methods for lasso model under different settings. Top row: computation time in second. Bottom row: relative computation time compared to AC. First column: Varying number of grid points $L$. Second column: Varying range of grid $\lambda_{\min}/\lambda_{\max}$. Third column: Varying number of grid points $L$ with fixed spacing.}
    \label{fig:5.1.1b}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{512.eps}    \caption{Comparing speed of screening methods for sparse logistic regression model under different settings. Top row: computation time in second. Bottom row: relative computation time compared to SSR. First column: varying number of features $p$. Second column: varying sample size $n$. Third column: varying signal strength $\beta_{\max}$. Fourth column: varying sparsity $s$.}
    \label{fig:5.1.2a}
\end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[width=0.82\textwidth]{512b.eps}    \caption{Comparing speed of screening methods for sparse logistic regression model under different settings. Top row: computation time in second. Bottom row: relative computation time compared to SSR. First column: Varying number of grid points $L$. Second column: Varying range of grid $\lambda_{\min}/\lambda_{\max}$. Third column: Varying number of grid points $L$ with fixed spacing.}
    \label{fig:5.1.2b}
\end{figure}

@@

\begin{table}[H]
\centering
\begin{tabular}{lllll}
\toprule
 & GENE & MNIST & GWAS & NYT \\
Screening method & n=536 & n=784 & n=313 & n=5,500 \\
       & p=17,322 & p=60,000 & p=660,496 & p=55,000 \\
\midrule
AC & 2.14 (0.03) & 6.82 (0.07) & 46.67 (0.13) & 59.20 (1.89) \\
SSR & 1.34 (0.01) & 5.20 (0.01) & 25.34 (0.08) & 35.40 (0.44) \\
SEDPP & 1.38 (0.01) & 8.40 (0.45) & 25.48 (0.05) & 53.69 (3.81) \\
HSSR & 1.21 (0.01) & 3.44 (0.07) & 23.85 (0.06) & 31.05 (0.80) \\
AHR & \textbf{0.71 (0.01)} & \textbf{1.15 (0.02)} & \textbf{8.80 (0.03)} & \textbf{12.71 (1.05)} \\
\bottomrule
\end{tabular}
\caption{Average computing time in second (standard error) for solving the lasso model}
\label{Tab:real}
\end{table}

AHR again outperforms other screening methods on all data sets, faster than other methods by factor ranging from 70\% to 630\%. Also note that the performance of SEDPP varies considerably for the MNIST and NYT data sets, where the response is randomly selected in each replication, indicating that the speed of SEDPP is sensitive to the data. AHR is built with SEDPP as the safe rule but does not share this issue of instability because it can adaptively adjust itself according the performance of the safe rule. Besides the computation time, we also look at a variaty of other statistics in the experiments, as discussed in the Appendix \ref{sec:ap}.

@@

\begin{table}[H]
\centering
\begin{tabular}{llllll}
\toprule
Screening method & AC\,\,\,\,\,\,\,\,\,    & SSR\,\,\,\,\,\,   & SEDPP & HSSR\,\,\,  & AHR\,\,\,\,\,\,  \\
\midrule
Time & 183.6 & 78.6 & 78.6 & 84.9 & 13.6 \\
Standard error & 0.43 & 0.65 & 0.20 & 0.48 & 0.15 \\
\bottomrule
\end{tabular}
\caption{Average computing time in minutes and standard error for the LVEF analysis (lasso-penalized linear regression).\label{Tab:lvef}}
\end{table}

@@

\begin{table}[H]
\centering
\begin{tabular}{llll}
\toprule
Screening method & SSR\,\,\,\,\,\,\,\,\,    & HSlores\,\,\,\,\,\,   & ASlores \\
\midrule
Time & 82.7 & 107.5 & 51.7 \\
Standard error & 0.10 & 0.67 & 0.17 \\
\bottomrule
\end{tabular}
\caption{Average computing time in minutes and standard error for the NYHA analysis (lasso-penalized logistic regression).\label{Tab:nyha}}
\end{table}

>>>>

This manuscript provides a brief review of existing feature screening rules in Section~\ref{sec:existing} as well as the following novel contributions:

\begin{itemize}
    \item A new feature screening framework (Sections~\ref{sec:asr}, \ref{sec:ahr}), which updates the reference for screening along the grid of $\lambda$ values on any combination of strong screening rules and safe screening rules, which can generate a family of more efficient and scalable screening rules, compared to previous screening rules that are either sequential or one-step.
    \item Three specific applications of this framework for fitting lasso penalized linear regression, lasso penalized logistic regression, and group lasso models (Sections~\ref{sec:method},~\ref{sec:logis},~\ref{sec:group}).
    \item An algorithm to determine when to update the reference $\lambda$ value for best efficiency (Section \ref{sec:adaptive}).
    \item Simulation studies (Section~\ref{sec:sim}) and real data analysis (Section~\ref{sec:real-data}) under a variety of scenarios, including testing data sets larger than memory, which showed that the proposed algorithms outperform other approaches in all cases, with substantial speed up.
    \item The new feature screening methods for lasso-penalized linear regression and logistic regression were implemented and added as new options to the publicly accessible \textbf{R} package \textbf{biglasso} \citep{zeng2017biglasso}, which was used to carry out the analyses in this manuscript.
\end{itemize}

@@

\subsection{Adaptive Safe Rules}
\label{sec:asr}

@@

\subsection{Adaptive Hybrid Rules}
\label{sec:ahr}

@@

\begin{algorithm}[ht]
  \label{alg:path}
    \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Pre}{Pre-compute}\SetKwInOut{Initialize}{Initialize}
    \SetAlgoLined
    \Input{$\{x_j\}_{j=1}^p, \{x_j^Ty\}_{j=1}^p,  \lambda_{0},...,\lambda_{L}$}
    \BlankLine
    \Initialize{reference = $\lambda_0$}

@@

This complexity analysis will only consider calculations with $O(np)$ complexity. The remaining calculations, those with $O(p)$ complexity for example, are not only of lower order, but also consume less memory and thus usually do no require out-of-core computing. Calculation of $\{x_j^Ty\}_{j=1}^p$ is needed only once for the whole solution path; we ignore this as well as its computational cost is the same across all approaches. We assume that strong rule screening is always powerful in the sense that its set of non-discarded features, $\mathcal{R}_{l+b}$, satisfies $|\mathcal{R}_{l+b}|=o(p) \approx 0$, but that safe rule screening, with corresponding set $\mathcal{S}_{l+b}$, may not be. We also assume that post-convergence check will usually pass in the first iteration and thus is only needed approximately once per $\lambda$. These two assumptions are supported by experiments from the original SSR paper \citep{Tibshirani2012} as well as our own investigations as discussed in the Appendix.

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{511.eps}    \caption{Comparing speed of screening methods for lasso model under different settings. Top row: computation time in second. Bottom row: relative computation time compared to AC. First column: varying number of features $p$. Second column: varying sample size $n$. Third column: varying signal strength $\beta_{\max}$. Fourth column: varying sparsity $s$.}
    \label{fig:5.1.1a}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.82\textwidth]{511b.eps}    \caption{Comparing speed of screening methods for lasso model under different settings. Top row: computation time in second. Bottom row: relative computation time compared to AC. First column: Varying number of grid points $L$. Second column: Varying range of grid $\lambda_{\min}/\lambda_{\max}$. Third column: Varying number of grid points $L$ with fixed spacing.}
    \label{fig:5.1.1b}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{512.eps}    \caption{Comparing speed of screening methods for sparse logistic regression model under different settings. Top row: computation time in second. Bottom row: relative computation time compared to SSR. First column: varying number of features $p$. Second column: varying sample size $n$. Third column: varying signal strength $\beta_{\max}$. Fourth column: varying sparsity $s$.}
    \label{fig:5.1.2a}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.82\textwidth]{512b.eps}    \caption{Comparing speed of screening methods for sparse logistic regression model under different settings. Top row: computation time in second. Bottom row: relative computation time compared to SSR. First column: Varying number of grid points $L$. Second column: Varying range of grid $\lambda_{\min}/\lambda_{\max}$. Third column: Varying number of grid points $L$ with fixed spacing.}
    \label{fig:5.1.2b}
\end{figure}

@@

\begin{table}[ht]
\centering
\begin{tabular}{lllll}
\toprule
 & GENE & MNIST & GWAS & NYT \\
Screening method & n=536 & n=784 & n=313 & n=5,500 \\
       & p=17,322 & p=60,000 & p=660,496 & p=55,000 \\
\midrule
AC & 2.14 (0.03) & 6.82 (0.07) & 46.67 (0.13) & 59.20 (1.89) \\
SSR & 1.34 (0.01) & 5.20 (0.01) & 25.34 (0.08) & 35.40 (0.44) \\
SEDPP & 1.38 (0.01) & 8.40 (0.45) & 25.48 (0.05) & 53.69 (3.81) \\
HSSR & 1.21 (0.01) & 3.44 (0.07) & 23.85 (0.06) & 31.05 (0.80) \\
AHR & \textbf{0.71 (0.01)} & \textbf{1.15 (0.02)} & \textbf{8.80 (0.03)} & \textbf{12.71 (1.05)} \\
\bottomrule
\end{tabular}
\caption{Average computing time in seconds (standard error) for solving the lasso model}
\label{Tab:real}
\end{table}

AHR again outperforms other screening methods on all data sets, faster than other methods by a factor ranging from 70\% to 630\%. Also note that the performance of SEDPP varies considerably for the MNIST and NYT data sets, where the response is randomly selected in each replication, indicating that the speed of SEDPP is sensitive to the data. AHR is built with SEDPP as the safe rule but does not share this issue of instability because it adaptively adjusts itself according the performance of the safe rule. A more detailed investigation of the algorithms' performance on these data sets including information on the number of rejections and KKT checks required is provided in the Appendix.

@@

\begin{table}[ht]
\centering
\begin{tabular}{llllll}
\toprule
Screening method & AC\,\,\,\,\,\,\,\,\,    & SSR\,\,\,\,\,\,   & SEDPP & HSSR\,\,\,  & AHR\,\,\,\,\,\,  \\
\midrule
Time & 183.6 & 78.6 & 78.6 & 84.9 & 13.6 \\
Standard error & 0.43 & 0.65 & 0.20 & 0.48 & 0.15 \\
\bottomrule
\end{tabular}
\caption{Average computing time in minutes and standard error for the LVEF analysis (lasso-penalized linear regression).\label{Tab:lvef}}
\end{table}

@@

\begin{table}[ht]
\centering
\begin{tabular}{llll}
\toprule
Screening method & SSR\,\,\,\,\,\,\,\,\,    & HSlores\,\,\,\,\,\,   & ASlores \\
\midrule
Time & 82.7 & 107.5 & 51.7 \\
Standard error & 0.10 & 0.67 & 0.17 \\
\bottomrule
\end{tabular}
\caption{Average computing time in minutes and standard error for the NYHA analysis (lasso-penalized logistic regression).\label{Tab:nyha}}
\end{table}
